Efficient implementation of penalized regression\\for genetic risk prediction

### 2019-02-09 paper/PRS.tex
save

{\color{red}
Applying our model to 350,000 individuals of the UK Biobank, we predict height with a larger correlation than with the best prediction of C+T [TODO CHIFFRES FROM TO], further demonstrating its scalibility and strong predictive power even for highly polygenic traits.
Moreover, using 150,000 individuals of the UK Biobank, we are able to predict breast cancer better than C+T, fitting PLR in a few minutes only.
This demonstrates the feasibility and relevance of using penalized regression for PRS computation when large individual-level datasets are available.

@@

Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRS are useful for genetic epidemiology when testing polygenicity of diseases and finding a common genetic contribution between two diseases \cite[]{purcell2009common}. Personalized medicine is another major application of PRS. Personalized medicine envisions to use PRS in screening campaigns in order to identify high-risk individuals for a given disease \cite[]{chatterjee2016developing}. As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}. 
{\color{red}
However, in order to be used in clinical settings, PRS should discriminate well enough between cases and controls. For screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact} [TODO: find prevalence].
}

@@

\begin{table}[htbp]
\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box. [TODO: number of SNPs in dataset]\label{tab:simus}}
\vspace*{0.5em}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
Numero of & \multirow{2}{*}{Dataset} & Sample size & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
scenario & & of training set & (number and location) & of effects & & model & \\
\hline
\hline
\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{ADD} & C+T \\
& & & 30 in all & & & & PLR \\
& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{COMP} & PLR3 \\
& & & 3000 in all & & & & (T-Trees) \\
\hline
\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{ADD} & C+T \\ 
& & & & & & & PLR\\
\hline
\multirow{5}{*}{3} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
& & 2000 & & & & & \\
& & 3000 & & & & & \\
& & 4000 & & & & & \\
& & 5000 & & & & & \\
\hline
\end{tabular}
\end{table}

@@

\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
\caption{Main comparison of C+T and PLR when simulating phenotypes with additive effects (scenario \textnumero1, model ``ADD'') [TODO: chnages for every plot]. Mean AUC over 100 simulations for PLR and the maximum AUC reported with ``C+T-max'' (clumping threshold at $r^2 > 0.2$). Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for a heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:main-AUC-logit}
\end{figure}

@@

\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
\caption{Comparison of methods in scenario \textnumero2 (using chromosome 6 only) for model ``ADD''. Thinner lines represents results in scenario \textnumero1. Mean AUC over 100 simulations for PLR and the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) as a function of the number and location of causal SNPs. Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for a heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-all-r2}
\end{figure}

\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
\caption{Comparison of methods when varying sample size in scenario \textnumero3 for model ``ADD'' with 300 causal SNPs sampled anywhere on the genome. Mean AUC over 100 simulations for the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the training size. Upper (lower) panels are presenting results for effects following a Gaussian (Laplace) distribution and left (right) panels are presenting results for a heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:main-AUC-ntrain}
\end{figure}

@@

Joint logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with PLR and 89.1\% with PLR3 as compared to 82.5\% with ``C+T-max'' (Table \ref{tab:results-celiac} and figure \ref{fig:celiac-roc}).
The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with PLR, PLR3 and ``C+T-max'', respectively.
Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360.
In terms of computation time, we show that PLR, while learning jointly on all SNPs at once and testing four different values for hyper-parameter $\alpha$, is almost as fast as the C+T method (190 vs 130 seconds), and PLR3 takes less than twice as long as PLR (296 vs 190 seconds). [TODO: reformuler SparSNP reponse]

@@

Our implementation of L1-penalized linear regression runs in less than one day for 350K individuals, achieving a correlation of more than 65.5\% with true height for both sexes [TODO: IN.. ajouter test set]. By comparison, the best C+T model achieves a correlation of 55\% for women and 56\% for men, and the GWAS part takes 1 hour. 
If using only the top 100,000 SNPs from a GWAS on the training set to train our L1-PLR, correlation between predicted and true heights drops at 0.634 for women and 0.643 for men.
Our L1-penalized logistic regression on breast cancer runs in 13 minutes for 150K women, achieving an AUC of 0.598 [TODO test set]. The best C+T model achieves an AUC of 0.589, and the GWAS part takes 15 hours.

\begin{table}[h]
\caption{[TODO deplace dfmax supmat] Results for the UK Biobank. The training/test sets for height (resp. breast cancer) are 350,000/24,131 (resp. 150,000/38,628). For PLR, dfmax is a parameter that controls the maximum number of variables allowed to enter the model before stopping fitting (default is 50,000 for each model corresponding to each validation set). Third row corresponds to fitting PLR on the top 100,000 SNPs from a GWAS on the training set. For height, $r$ (correlation with true height) is reported within women / men; for breast cancer, AUC is reported.\label{tab:results-UKBB}}
\vspace*{0.5em}
\centering
\begin{tabular}{|l|l|c|c|c|}
  \hline
Trait & Method & $r$ (women / men) or AUC & \# predictors & Execution time \\
  \hline
height & PLR (dfmax = 50e3) & 0.656 / 0.657 & 115,997 & 21h \\ 
height & PLR (dfmax = Inf)  & 0.657 / 0.658 & 148,341 & 61h \\ 
height & PLR (100K top SNPs, dfmax = Inf)  & 0.634 / 0.643 & 70,162 & 48.5h \\ 
height & C+T-max & 0.549 / 0.561 & 45,570 & 69m \\ 
breast cancer & PLR & 0.598 & 2653 & 13m \\ 
breast cancer & C+T-max & 0.589 & 21 & 15h \\
  \hline
\end{tabular}
\end{table}

@@

When there are many small effects and a small sample size, PLR performs worse than (the best result for) C+T. For example, this underpowered situation occurs when there are many causal variants (3K) to find among many typed variants (280K) and a small sample size (6K). 
In such underpowered scenarios, it is difficult to detect true causal variants, which makes PLR too conservative whereas the best strategy is to include nearly all SNPs \cite[]{purcell2009common}.
When increasing sample size (scenario \textnumero3), PLR achieves higher predictive performance than C+T and the benefits of using PLR over C+T increase with an increasing sample size (Figure \ref{fig:main-AUC-ntrain}). Moreover, when decreasing the search space (total number of candidate SNPs) in scenario \textnumero2, we increase the proportion of causal variants and we virtually increase the sample size \cite[]{dudbridge2013power}. In this scenario \textnumero2, even when there are small effects and a high polygenicity (3000 causal variants out of 18,941), PLR gets a large increase in predictive performance, now consistently higher than C+T (Figure \ref{fig:supp-AUC-all-r2}).
}

@@

Memory requirements are tightly linked to computation time. Indeed, active variables are accessed in memory thanks to memory-mapping when they are used \cite[]{prive2017efficient}. 
If there are too many variables accessed, the operating system (OS) must free some memory for new coming variables. 
There should never be any memory issue thanks to memory-mapping. Yet, if there are too many active variables, the OS would regularly swap memory between disk and RAM, severely slowing down computations. 
%If this happens for very large datasets, you can try to use less processing cores (as active sets tend to have non-overlapping variables because of LD) or buy more RAM. 
%Another solution would be to reduce the `n.abort' parameter (default value is 10) in order to stop fitting the regularization path even sooner. 
[TODO: add doc function]
%This would avoid fitting models with a large number of variables without changing the solution since regularization paths for very large sample sizes are very smooth (Figure \ref{fig:supp-path}).
%We tested processing each chromosome separately and then combining the models later but this did not perform as well as fitting all SNPs at once (data not shown). 
A possible approach to reduce computational burden is to apply penalized regression on a subset of SNPs by prioritizing SNPs using univariate tests (GWAS computed from the same dataset). Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm in this paper. 
Indeed, when using only the 100K most significantly associated SNPs, this reduces correlation between predicted and true heights from 0.657/0.658 to 0.634/0.643 within women/men, while decreasing computation time by 20\% only (Table \ref{tab:results-UKBB}). 
A key advantage of our implementation of PLR is that prior filtering of variables is no more required for computational feasibility.

>>>>

{\color{red}
Applying our model to 350,000 individuals of the UK Biobank, we predict height with a larger correlation than with the best prediction of C+T (\textasciitilde65\% instead of \textasciitilde55\%), further demonstrating its scalibility and strong predictive power, even for highly polygenic traits.
Moreover, using 150,000 individuals of the UK Biobank, we are able to predict breast cancer better than C+T, fitting PLR in a few minutes only.
This demonstrates the feasibility and relevance of using penalized regression for PRS computation when large individual-level datasets are available.

@@

Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRS are useful for genetic epidemiology when testing polygenicity of diseases and finding a common genetic contribution between two diseases \cite[]{purcell2009common}. Personalized medicine is another major application of PRS. Personalized medicine envisions to use PRS in screening campaigns in order to identify high-risk individuals for a given disease \cite[]{chatterjee2016developing}. As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}. 
{\color{red}
However, in order to be used in clinical settings, PRS should discriminate well enough between cases and controls. For screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that, for a 10\% disease prevalence, the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.
}

@@

\begin{table}[htbp]
\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box. \label{tab:simus}}
\vspace*{0.5em}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
Numero of & Dataset & Sample size & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
scenario & (number of SNPs) & of training set & (number and location) & of effects & & model & \\
\hline
\hline
\multirow{4}{*}{1} & \multirow{3}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{ADD} & C+T \\
& & & 30 in all & & & & PLR \\
& (281,122 SNPs) & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{COMP} & PLR3 \\
& & & 3000 in all & & & & (T-Trees) \\
\hline
\multirow{2}{*}{2} & Chromosome 6 only & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{ADD} & C+T \\ 
& (18,941 SNPs) & & & & & & PLR\\
\hline
\multirow{5}{*}{3} & \multirow{4}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
& \multirow{4}{*}{(281,122 SNPs)} & 2000 & & & & & \\
& & 3000 & & & & & \\
& & 4000 & & & & & \\
& & 5000 & & & & & \\
\hline
\end{tabular}
\end{table}

@@

\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
\caption{Main comparison of C+T and PLR when simulating phenotypes with additive effects (scenario \textnumero1, model ``ADD''). Mean AUC over 100 simulations for PLR and the maximum AUC reported with ``C+T-max'' (clumping threshold at $r^2 > 0.2$). Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for a heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:main-AUC-logit}
\end{figure}

@@

\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-chr6-all-r2}}
\caption{Comparison of methods when simulating phenotypes with additive effects and using chromosome 6 only (scenario \textnumero2). Thinner lines represents results in scenario \textnumero1. Mean AUC over 100 simulations for PLR and the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) as a function of the number and location of causal SNPs. Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for a heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-all-r2}
\end{figure}

\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
\caption{Comparison of methods when simulating 300 causal SNPs with additive effects and when varying sample size (scenario \textnumero3). Mean AUC over 100 simulations for the maximum values of C+T for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and PLR as a function of the training size. Upper (lower) panels are presenting results for effects following a Gaussian (Laplace) distribution and left (right) panels are presenting results for a heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:main-AUC-ntrain}
\end{figure}

@@

Joint logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with PLR and 89.1\% with PLR3 as compared to 82.5\% with ``C+T-max'' (Table \ref{tab:results-celiac} and figure \ref{fig:celiac-roc}).
The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with PLR, PLR3 and ``C+T-max'', respectively.
Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360.
In terms of computation time, we show that PLR, while learning jointly on all SNPs at once and testing four different values for hyper-parameter $\alpha$, is almost as fast as the C+T method (190 vs 130 seconds), and PLR3 takes less than twice as long as PLR (296 vs 190 seconds).

@@

Our implementation of L1-penalized linear regression runs in less than one day for 350K individuals (training set), achieving a correlation of more than 65.5\% with true height for both sexes in the remaining 24K individuals (test set). By comparison, the best C+T model achieves a correlation of 55\% for women and 56\% for men (in the test set), and the GWAS part takes one hour (for the training set). 
If using only the top 100,000 SNPs from a GWAS on the training set to fit our L1-PLR, correlation between predicted and true heights drops at 0.634 for women and 0.643 for men.
Our L1-penalized logistic regression on breast cancer runs in 13 minutes for 150K women, achieving an AUC of 0.598 in the remaining 39K women, while the best C+T model achieves an AUC of 0.589, and the GWAS part takes 15 hours.

\begin{table}[h]
\caption{Results for the UK Biobank. The sizes of training / test sets for height (resp.\ breast cancer) are 350,000 / 24,131 (resp.\ 150,000 / 38,628). For height, $r$ (correlation between predicted and true heights) is reported within women / men separately; for breast cancer, AUC is reported.\label{tab:results-UKBB}}
\vspace*{0.5em}
\centering
\begin{tabular}{|l|l|c|c|c|}
  \hline
Trait & Method & $r$ (women / men) & \# predictors & Execution time \\
  \hline
height & PLR & 0.656 / 0.657 & 115,997 & 21h \\ 
height & C+T-max & 0.549 / 0.561 & 45,570 & 69m \\ 
  \hline
  \hline
Trait & Method & AUC & \# predictors & Execution time \\
  \hline
breast cancer & PLR & 0.598 & 2653 & 13m \\ 
breast cancer & C+T-max & 0.589 & 21 & 15h \\
  \hline
\end{tabular}
\end{table}

@@

When there are many small effects and a small sample size, PLR performs worse than (the best result for) C+T. For example, this situation occurs when there are many causal variants (3K) to distinguish among many typed variants (280K) while using a small sample size (6K). 
In such underpowered scenarios, it is difficult to detect true causal variants, which makes PLR too conservative whereas the best strategy is to include nearly all SNPs \cite[]{purcell2009common}.
When increasing sample size (scenario \textnumero3), PLR achieves higher predictive performance than C+T and the benefits of using PLR over C+T increase with an increasing sample size (Figure \ref{fig:main-AUC-ntrain}). Moreover, when decreasing the search space (total number of candidate SNPs) in scenario \textnumero2, we increase the proportion of causal variants and we virtually increase the sample size \cite[]{dudbridge2013power}. In this scenario \textnumero2, even when there are small effects and a high polygenicity (3000 causal variants out of 18,941), PLR gets a large increase in predictive performance, now consistently higher than C+T (Figure \ref{fig:supp-AUC-all-r2}).
}

@@

Memory requirements are tightly linked to computation time. Indeed, active variables are accessed in memory thanks to memory-mapping when they are used \cite[]{prive2017efficient}. 
When there is not enough memory left, the operating system (OS) frees some memory for new coming variables. 
There should never be any memory issue thanks to memory-mapping. Yet, if there are too many active variables, the OS would regularly swap memory between disk and RAM, severely slowing down computations. 
%If this happens for very large datasets, you can try to use less processing cores (as active sets tend to have non-overlapping variables because of LD) or buy more RAM. 
%Another solution would be to reduce the `n.abort' parameter (default value is 10) in order to stop fitting the regularization path even sooner. 
%This would avoid fitting models with a large number of variables without changing the solution since regularization paths for very large sample sizes are very smooth (Figure \ref{fig:supp-path}).
%We tested processing each chromosome separately and then combining the models later but this did not perform as well as fitting all SNPs at once (data not shown). 
A possible approach to reduce computational burden is to apply penalized regression on a subset of SNPs by prioritizing SNPs using univariate tests (GWAS computed from the same dataset). Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm in this paper. 
Indeed, when using only the 100K most significantly associated SNPs, this reduces correlation between predicted and true heights from 0.657/0.658 to 0.634/0.643 within women/men, while decreasing computation time by 20\% only (Table \ref{tab:results-UKBB}). 
A key advantage of our implementation of PLR is that prior filtering of variables is no more required for computational feasibility.
